Learning CNN filters from user-drawn image markers for coconut-tree image classification

### 2020-07-24 main.tex
Updates from Overleaf

The design of CNN models without a human as part of the training loop leaves several questions unanswered: (1) How to find a useful and simplified model for a given classification problem? (2) How to train that model from a minimum number of annotated images?  (3) Can the user explain the decisions of the model? (4) Can the model improve from label corrections? The first question requires human knowledge about CNN modeling and the problem of interest. The second one requires to reduce the human effort to train a CNN model. The third issue is related to human understanding. It may explore data visualization techniques to explain the model's decisions and to guide the user in the design of the model. The fourth question is also essential during training, and it is related to user control over the process. They all lead to the importance of involving human experts during the machine learning process. \textcolor{red}{Explainable artificial intelligence (XAI) has appeared to address such problems~\cite{rauber2016visualizing,arrietaInfoFusion2020}. However, the importance of human participation during the training process has called little attention~\cite{AngelovNN2020}. Exceptions rely on visual analytics techniques~\cite{GarciaIJCNN2019,SpinnerTVCG2020}, and we present here a first work in which the user has direct influence over the quality of the filters.}

First, we conceptually divide a CNN into convolutional layers for feature extraction and fully connected layers for feature space reduction and classification. Each convolutional layer contains a filter bank, an activation function, and alternative operations (e.g., pooling, batch normalization).  As the number of convolutional layers increases (deeper is the model), and higher is the number of annotated images required to train the model by backpropagation.

To reduce the need for large annotated training sets by exploiting the user knowledge, we present a method, called \textcolor{red}{\emph{Feature Learning from Image Markers} (FLIM)}, that needs a minimal set of images to learn the filters of each convolutional layer. The user selects the number of convolutional layers, their filter sizes, \textcolor{red}{a few images from a training set}, and draws markers in image regions that best discriminate the classes. \textcolor{red}{By clustering patches extracted from marker pixels at the input of each convolutional layer, FLIM finds filters that enhance those regions at the output of the convolutional layers.} The CNN architecture could also be optimized based on visual analytics~\cite{SpinnerTVCG2020}.

We demonstrate the advantages of this interactive technique on the binary image classification of coconut-tree images against  VGG-16~\cite{simonyan2014very}. First, the user can better understand and control the training process by observing the effectiveness of marker selection.  Second, very few images per class (e.g., less than five) seem to be enough. Third, by eliminating the optimization of the convolutional layers by backpropagation, FLIM can reduce the number of annotated images to train the fully connected layers. Fourth, FLIM is application-independent, and so it might be useful for other image classification problems.

@@

\textcolor{red}{A CNN model should enhance image regions that best discriminate classes through a sequence of convolutional layers. This section shows how the identification of such regions by a simple user action can make that sequence an effective feature extractor. The method is named \emph{Feature Learning from Image Markers} (FLIM). In the feature space that results from convolutional layers,} it is common to assume that the classes are piecewise linearly separable and apply fully connected layers -- a Multi-Layer Perceptron (MLP) classifier -- for feature space reduction and classification. The training of a CNN usually relies on weight optimization using backpropagation, which requires a higher number of annotated images as deep as the model is. In FLIM, we eliminate the need for backpropagation to train convolutional layers by finding a set ${\cal F}$ of filters that can detect discriminant regions from all classes.

The user selects a few training images to compose a very small dataset $\D$ and draws labeled markers in image regions that best discriminate the classes. The convolution between an image $I$ and a filter $F\in {\cal F}$ generates positive activations in regions whose local patterns are detected by $F$. We wish then to estimate the weights of $F$ such that those local patterns are characteristic of one given class. For sake of clarity, one must interpret the convolution operation at a pixel $p$ of $I$ as the inner product $\vect(P_I(p)) \cdot \vect(F)$, where $P_I(p)$ is a $k\times k$ patch with $m$ channels around $p$, $F$ is a filter with shape $k\times k \times m$, and $\vect$ is the vectorization operation. A filter $F$ can discriminate a class $i \in \{1,2,\ldots,c\}$ among $c$ classes when it generates positive results for local patterns $\vect(P_I(p))$ of class $i$ and negative results for patterns from other classes. That is, $\vect(F)$ is the normal vector of a hyperplane in $\R^{k\times k\times m}$, which detects the patterns of class $i$ at its positive side. We wish to estimate ${\cal F}$ such that its filters will detect patterns from all classes in different positions $p$.

>>>>

The design of CNN models without a human as part of the training loop leaves several questions unanswered: (1) How to find a useful and simplified model for a given classification problem? (2) How to train that model from a minimum number of annotated images?  (3) Can the user explain the decisions of the model? (4) Can the model improve from label corrections? The first question requires human knowledge about CNN modeling and the problem of interest. The second one requires to reduce the human effort to train a CNN model. The third issue is related to human understanding. It may explore visual analytics to explain the model's decisions and to guide the user in the design of the model~\cite{RauberInfoVis2018,GarciaIJCNN2019,SpinnerTVCG2020}. The fourth question is also essential during training, and it is related to user control over the process. They all lead to the importance of involving human experts during the machine learning process. \textcolor{red}{Explainable artificial intelligence (XAI) has appeared to address such problems~\cite{rauber2016visualizing,arrietaInfoFusion2020}. However, the importance of human participation during the training process has called little attention~\cite{AngelovNN2020}. Exceptions rely on visual analytics~\cite{RauberInfoVis2018,GarciaIJCNN2019,SpinnerTVCG2020}, and we present here a first work in which the user has direct influence over the quality of the filters.}

First, we conceptually divide a CNN into convolutional layers for feature extraction and fully connected layers for feature space reduction and classification. Each convolutional layer contains a filter bank, an activation function, and alternative operations (e.g., pooling, batch normalization).  As the number of convolutional layers increases (deeper is the model), higher is the number of annotated images required to train the model by backpropagation.

To reduce the need for large annotated training sets by exploiting the user knowledge, we present a method, called \textcolor{red}{\emph{Feature Learning from Image Markers} (FLIM)}, that needs a minimal set of images to learn the filters of each convolutional layer. The user selects the number of convolutional layers, their filter sizes, \textcolor{red}{a few images from a training set}, and draws markers in image regions that best discriminate the classes. \textcolor{red}{By clustering patches extracted from marker pixels at the input of each convolutional layer, FLIM finds filters that enhance those regions at the output of the convolutional layers. The remaining training images can then be used to train the fully connected layers.} The CNN architecture could also be optimized based on visual analytics~\cite{SpinnerTVCG2020}, but we have not explored this option yet.

We demonstrate the advantages of this interactive technique on binary classification of coconut-tree images against  VGG-16~\cite{simonyan2014very}. First, the user can better understand and control the training process by observing the effectiveness of marker selection.  Second, very few images per class (e.g., less than five) seem to be enough. Third, by eliminating the optimization of the convolutional layers by backpropagation, FLIM can reduce the number of annotated images to train the fully connected layers. Fourth, FLIM is application-independent, and so it might be useful for other image classification problems.

@@

\textcolor{red}{A CNN model should enhance image regions that best discriminate classes through a sequence of convolutional layers. This section shows how the identification of such regions by a simple user action can make that sequence an effective feature extractor. The method is named \emph{Feature Learning from Image Markers} (FLIM). In the feature space that results from convolutional layers,} it is common to assume that the classes are piecewise linearly separable and apply fully connected layers -- Multi-Layer Perceptrons (MLP) -- for feature space reduction and classification. The training of a CNN usually relies on weight optimization using backpropagation, which requires a higher number of annotated images as deep as the model is. In FLIM, we eliminate the need for backpropagation to train convolutional layers by finding a set ${\cal F}$ of filters that can \textcolor{red}{enhance} discriminant regions from all classes.

The user selects a few \textcolor{red}{images from a training set} to compose a very small dataset $\D$ and draws labeled markers in image regions that best discriminate the classes. The convolution between an image $I$ and a filter $F\in {\cal F}$ \textcolor{red}{ enhances} regions whose local patterns are detected by $F$. We wish then to estimate the weights of $F$ such that those local patterns are characteristic of one given class. For sake of clarity, one must interpret the convolution operation at a pixel $p$ of $I$ as the inner product $\vect(P_I(p)) \cdot \vect(F)$, where $P_I(p)$ is a $k\times k$ patch with $m$ channels around $p$, $F$ is a filter with shape $k\times k \times m$, and $\vect$ is the vectorization operation. A filter $F$ can discriminate a class $i \in \{1,2,\ldots,c\}$ among $c$ classes when it generates positive results for local patterns $\vect(P_I(p))$ of class $i$ and negative results for patterns from other classes. That is, $\vect(F)$ is the normal vector of a hyperplane in $\R^{k\times k\times m}$, which detects the patterns of class $i$ at its positive side. We wish to estimate ${\cal F}$ such that its filters will detect patterns from all classes in different positions $p$.
